Doublet splitting of fusion alpha-particle driven ion cyclotron emission

### 2022-07-20 manuscript.tex
added refs to Liu and Ochoukov, multiplet splitting, drift resonances

To better enable ICE as a diagnostic for confined fast ions, it is necessary to
further develop our understanding with a range of tools. In this work, I present
the growth rates in 2D wavenumber space of the MCI under the conditions of
outer mid-plane of the JET PTE\cite{Cottrell1993} Pulse No. 26148. These growth
rates are obtained from a dispersion relation solver,
LinearMaxwellVlasov.jl\cite{LinearMaxwellVlasov.jl}, written in the Julia programming
language\cite{Bezanson2017}. These results narrow the gap between analytical
theory and expensive HPC simulations and can be run readily on desktop computers
in a matter of minutes, further enabling research into this topic. Note that
this study focusses on edge-ICE, named for the region of plasma from which the signal
arises, as opposed to core-ICE\cite{Ochoukov2018}.

@@

Doublet splitting can be seen in Fig. \ref{fig:2D_imag_physicalpitch}, which is the
counterpart figure to Fig. \ref{fig:2D_F12_physicalpitch} as Fig.
\ref{fig:2D_imag_zeropitch} is to Fig. \ref{fig:2D_F12_zeropitch},
at e.g. the $12^{th}$ harmonic  near $(k_\parallel \simeq 0\; \Omega_{i}/V_A,
k_\bot \simeq 13\; \Omega_{i}/V_A)$, where $V_A$ is the Alfv{\'e}n speed.
Other possible mechanisms for doublet splitting could be grad-B or curvature
drifts\cite{Cottrell1993,Fulop1998}. Qualitatively, scattered points in Fig.
\ref{fig:2D_F12_physicalpitch}(a) resembles the experimental observations given
by the black trace in that the
signal below the $7^{th}$ cyclotron harmonics is lower and flatter with well
defined peaks and the signal beyond the $7^{th}$ harmonic is a step-change
larger and the harmonics are less easily identifiable. These features cannot be
be recreated with a 1D calculation for these distributions functions. Note that
Fig. \ref{fig:2D_imag_physicalpitch} shows these two different regimes more
clearly. The lower frequency behaviour, where $k_\perp < 7 \Omega_i/V_A$, is
characterised by narrow regions of instability near Doppler shifted cyclotron
harmonics that are stable when $k_\parallel=0$ and unstable at larger
negative parallel wavenumbers. The higher frequency behaviour in the region
$\omega\geq 7\Omega_i$ is dominated by broader regions of instability in
frequency and wavenumber near to Doppler shifted cyclotron harmonics, which have
a complex structure in wavenumber space. Modes numbering 1, 2, 6, 7, 8, 12, and
13 are stable when propagating exactly perpendicular whilst obliquely
propagating modes remain unstable. The magnitude of the Doppler shift of
the growing modes can reach $\approx \Omega_i$ for large propagation angles at 
lower frequencies and also for
oblique angles at higher frequencies; see, for example, the mode
associated with the $15^{th}$ cyclotron harmonic near $k_\parallel \simeq
-\frac{3}{4} \Omega_i/V_A, k_\perp \simeq 14 \Omega_i/V_A$, i.e. a propagation
angle of $93^\circ$. Furthermore, the breadth of the harmonics can be
attributed to the contributions from multiple propagation angles even though the
lower harmonics are largely stable for forward travelling and perpendicular
propagation angles.

@@

>>>>

To better enable ICE as a diagnostic for confined fast ions, it is necessary to
further develop our understanding with a range of tools. In this work, I present
the growth rates in 2D wavenumber space of the MCI under the conditions of
outer mid-plane of the JET PTE\cite{Cottrell1993} Pulse No. 26148. These growth
rates are obtained from a dispersion relation solver,
LinearMaxwellVlasov.jl\cite{LinearMaxwellVlasov.jl}, written in the Julia programming
language\cite{Bezanson2017}. These results narrow the gap between analytical
theory and expensive HPC simulations and can be run readily on desktop computers
in a matter of minutes, further enabling research into this topic. Note that
this study focusses on edge-ICE, named for the region of plasma from which the signal
arises, as opposed to core-ICE\cite{Ochoukov2018}. Recent observations made with
higher resolution equipment than those referred to on JET have
shown multiplet splitting in NBI-driven spectra; from the core on
ASDEX-Upgrade\cite{Ochoukov2019} and edge on EAST\cite{Lui2020}.

@@

Doublet splitting can be seen in Fig. \ref{fig:2D_imag_physicalpitch}, which is the
counterpart figure to Fig. \ref{fig:2D_F12_physicalpitch} as Fig.
\ref{fig:2D_imag_zeropitch} is to Fig. \ref{fig:2D_F12_zeropitch},
at e.g. the $12^{th}$ harmonic  near $(k_\parallel \simeq 0\; \Omega_{i}/V_A,
k_\bot \simeq 13\; \Omega_{i}/V_A)$, where $V_A$ is the Alfv{\'e}n speed.
Other possible mechanisms for doublet splitting could be grad-B or curvature
drifts\cite{Cottrell1993,Fulop1998} or potentially any mechanism that is capable
of modifying the wave-particle resonance condition.
Qualitatively, scattered points in Fig.
\ref{fig:2D_F12_physicalpitch}(a) resembles the experimental observations given
by the black trace in that the
signal below the $7^{th}$ cyclotron harmonics is lower and flatter with well
defined peaks and the signal beyond the $7^{th}$ harmonic is a step-change
larger and the harmonics are less easily identifiable. These features cannot be
be recreated with a 1D calculation for these distributions functions. Note that
Fig. \ref{fig:2D_imag_physicalpitch} shows these two different regimes more
clearly. The lower frequency behaviour, where $k_\perp < 7 \Omega_i/V_A$, is
characterised by narrow regions of instability near Doppler shifted cyclotron
harmonics that are stable when $k_\parallel=0$ and unstable at larger
negative parallel wavenumbers. The higher frequency behaviour in the region
$\omega\geq 7\Omega_i$ is dominated by broader regions of instability in
frequency and wavenumber near to Doppler shifted cyclotron harmonics, which have
a complex structure in wavenumber space. Modes numbering 1, 2, 6, 7, 8, 12, and
13 are stable when propagating exactly perpendicular whilst obliquely
propagating modes remain unstable. The magnitude of the Doppler shift of
the growing modes can reach $\approx \Omega_i$ for large propagation angles at 
lower frequencies and also for
oblique angles at higher frequencies; see, for example, the mode
associated with the $15^{th}$ cyclotron harmonic near $k_\parallel \simeq
-\frac{3}{4} \Omega_i/V_A, k_\perp \simeq 14 \Omega_i/V_A$, i.e. a propagation
angle of $93^\circ$. Furthermore, the breadth of the harmonics can be
attributed to the contributions from multiple propagation angles even though the
lower harmonics are largely stable for forward travelling and perpendicular
propagation angles.

@@

\bibitem{Ochoukov2019}
Ochoukov R \textit{et al} 2019 \textit{Nucl. Fusion} \textbf{59} 086032

\bibitem{Lui2020}
Liu L \text{et al} 2020 \textit{Nucl. Fusion} \textbf{60} 044002
